---
title: "Estimation on 10x data using loom and pagoda2"
output: html_notebook
---

Here we show an example of how [loom-annotated](http://velocyto.org/velocyto.py/tutorial/index.html#running-the-cli) matrices of a 10x dataset can be loaded and analyzed in R using velocyto.R and [pagoda2](https://github.com/hms-dbmi/pagoda2).


## Data loading

Load the velocyto package:
```{r}
library(velocyto.R)
```


Load loom data
```{r}
# you can download 10X43_1.loom from the following URL: http://pklab.med.harvard.edu/velocyto/DG1/10X43_1.loom
ldat <- read.loom.matrices("10X43_1.loom")
```

## Normalize and cluster cells using pagoda2
Using spliced expression matrix as input to pagoda2.
```{r}
emat <- ldat$spliced

# this dataset has already been pre-filtered, but this is where one woudl do some filtering
emat <- emat[,colSums(emat)>=1e3]
```

## Pagoda2 processing

[Pagoda2](https://github.com/hms-dbmi/pagoda2) is used to generate cell embedding, cell clustering, as well as a more accurate cell-cell distance matrix. You can alternatively generate those using other tools, such as Seurat2, etc.

Create pagoda2 object, adjust variance:
```{r fig.height=3,fig.width=6}
library(pagoda2)
r <- Pagoda2$new(emat,modelType='plain',trim=10,log.scale=T)
r$adjustVariance(plot=T,do.par=T,gam.k=10)
```

Run basic analysis steps to generate cell embedding and clustering, visualize:
```{r}
r$calculatePcaReduction(nPcs=100,n.odgenes=3e3,maxit=300)
r$makeKnnGraph(k=30,type='PCA',center=T,distance='cosine');
r$getKnnClusters(method=multilevel.community,type='PCA',name='multilevel')
r$getEmbedding(type='PCA',embeddingType='tSNE',perplexity=50,verbose=T)
```

Plot embedding, labeling clusters (left) and "Xist" expression (which separates the male and female )
```{r fig.height=4,fig.width=8}
par(mfrow=c(1,2))
r$plotEmbedding(type='PCA',embeddingType='tSNE',show.legend=F,mark.clusters=T,min.group.size=10,shuffle.colors=F,mark.cluster.cex=1,alpha=0.3,main='cell clusters')
r$plotEmbedding(type='PCA',embeddingType='tSNE',colors=r$counts[,"Xist"],main='Xist')  

```

## Velocity estimation

Prepare matrices and clustering data:
```{r}
emat <- ldat$spliced; nmat <- ldat$unspliced
emat <- emat[,rownames(r$counts)]; nmat <- nmat[,rownames(r$counts)]; # restrict to cells that passed p2 filter
# take cluster labels
cluster.label <- r$clusters$PCA[[1]]
cell.colors <- pagoda2:::fac2col(cluster.label)
# take embedding
emb <- r$embeddings$PCA$tSNE
```


In addition to clustering and the t-SNE embedding, from the p2 processing we will also take a cell-cell distance, which will be better than the default whole-transcriptome correlation distance that velocyto.R would normally use.
```{r}
cell.dist <- as.dist(1-armaCor(t(r$reductions$PCA)))
```

Filter genes based on the minimum average expresion magnitude (in at least one of the clusters), output total number of resulting valid genes:
```{r}
emat <- filter.genes.by.cluster.expression(emat,cluster.label,min.max.cluster.average = 0.5)
nmat <- filter.genes.by.cluster.expression(nmat,cluster.label,min.max.cluster.average = 0.05)
length(intersect(rownames(emat),rownames(emat)))
```

Estimate RNA velocity (using gene-relative model with k=20 cell kNN pooling and using top/bottom 2% quantiles for gamma fit):
```{r}
fit.quantile <- 0.02
rvel.cd <- gene.relative.velocity.estimates(emat,nmat,deltaT=1,kCells=20,cell.dist=cell.dist,fit.quantile=fit.quantile)
```


Visualize velocity on the t-SNE embedding, using velocity vector fields:
```{r fig.height=5,fig.width=5}
show.velocity.on.embedding.cor(emb,rvel.cd,n=300,scale='sqrt',cell.colors=ac(cell.colors,alpha=0.5),cex=0.8,arrow.scale=5,show.grid.flow=TRUE,min.grid.cell.mass=0.5,grid.n=40,arrow.lwd=1,do.par=F,cell.border.alpha = 0.1)
```


Visualize a fit for a particular gene (we reuse rvel.cd to save on calcualtions here):
```{r fig.height=2.3,fig.width=8}
gene <- "Nfib"
gene.relative.velocity.estimates(emat,nmat,deltaT=1,kCells = 20,kGenes=1,fit.quantile=fit.quantile,cell.emb=emb,cell.colors=cell.colors,cell.dist=cell.dist,show.gene=gene,old.fit=rvel.cd,do.par=T)
```
